Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
edited simplify alg

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and all such instances are independent).
Given a node table $\Nt$, the function $\taddrow{\Nt}{t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddrow{\Et}{\ell, r, p, c}$
adds a new edge $(\ell\mbox{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
pybind11 \citep[version 2.2.1][]{pybind11},
and GCC (version 4.8.5).
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

Here we describe the simplification algorithm. The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represents the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval. The per-interval
segments are assigned to input IDs via the $\Al$ mapping, which is the main state required
by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node ID mapped to input
node ID $u$ over a set of intervals. For each input node, we need to find the process
the resulting set of ancestral segments to find overlaps (and hence coalescences). We
do this using a $\priorityq{}$ type. This priority queue maintains a list of ancestral
segments in sorted order, such that the segements with the smallest left coordinate
are at the start of the list. Given a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}{x}$
returns (and removes) the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

A step-by-step explanation of the algorithm follows the algorithm listing.

@@

\algstep{S1.}{Select parent edges.}{
    Set $E \leftarrow \{e: \attrparent{e} = u \mbox{ and } e \in \Et_I\}$.
    Set $v \leftarrow -1$.
}

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
    $\pqpush{Q}{y}$.
}

@@

\algstep{S5.}{New output node.}{
    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
}

@@

The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
state for the samples in $S$. For each input sample with node ID $u$, we
add a row in the output node table $\Nt_O$, and create a mapping for this
new node $v$. After this initial state has been created, we then allocate our
priority queue $Q$, and set the current input ID $u$ to $0$.

The algorithm steps back up through the pedigree and steps \algref{S1}--\algref{S8}
consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
In \algref{S1} we first find all edges in the input where the parent is equal to
our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
Then, in \algref{S2} we find all intersections between these edges and the
mapped ancestry segments for their child nodes. For each of these intersections we
create a new segment $y$ and insert it into the priority queue.

After finding all overlaps and filling the priority queue with the resulting ancestry
segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
loop over the segments in the queue, considering each in turn until the queue is empty.
Step \algref{S3} builds a list $X$ of all the segments that have the
minimum left coordinate (along with some bookkeeping to keep track of where our next
right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
this means that we have no overlapping ancestral segments at this coordinate and
we proceed to \algref{S4}.

As no overlaps occur, there is no change in the mapping between input and output IDs
and so the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
However, there is a complication caused by other segments which may intersect with $x$,
which have left coordinate greater then $\attrleft{x}$. If $x$ intersects with this value,
then there are other overlaps we must consider. Therefore, we change the left coordinate
of $x$ to this coordinate and reinsert it into $Q$ where it can be processed again
later. After doing this, we proceed to \algref{S7}, update the ancestry mapping for
the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.

On the other hand, if there is more than one segment in $X$, we know that an
overlap among the ancestral segments has occured and we must update the output
nodes and edges accordingly. In step \algref{S5} we first allocate a new output
node $v$ for $u$, if it has not been done already (we may have many overlapping
segments along the genome). Step \algref{S6} then continues by creating a new
ancestry segment $\alpha$ mapping the current interval to the output node $v$.
We then iterate over all segments in $X$, adding the appropriate output edges
and editing and reinserting the ancestry segment $x$ into the priority queue if
necessary, as before.

>>>>

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and all such instances are independent).
Given a node table $\Nt$, the function $\taddrow{\Nt}{t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddrow{\Et}{\ell, r, p, c}$
adds a new edge $(\ell\text{eft}, r\text{ight}, p\text{arent}, c\text{hild})$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|-1)$.
A step-by-step explanation follows the pseudocode.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).
Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
pybind11 \citep[version 2.2.1][]{pybind11},
and GCC (version 4.8.5).
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

Here we provide a concrete implementation of the simplification algorithm.
The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
These output node IDs correspond to colors in the ``paint pot'' analogy.
The per-interval segments are assigned to input IDs via the $\Al$ mapping,
which is the main state required by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node IDs mapped to input
node ID $u$ over a set of intervals.
For each input node, we need to process the ancestral segments
to find overlaps (and hence coalescences), and to update the $\Al$.
We do this using a $\priorityq{}$ type, which maintains a list of ancestral
segments sorted by left coordinate. For a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
removes and returns the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

@@

\algstep{S1.}{Select parent edges.}{
    Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S8}.
}

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrleft{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
    $\pqpush{Q}{y}$.
}

@@

\algstep{S5.}{Overlap; new output node.}{
    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
}

@@

The algorithm begins in \algref{S0} by allocating an empty list to store ancestral
segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
state for the samples in $S$. For each input sample with node ID $u$, we
add a row in the output node table $\Nt_O$, and create a mapping for this
new node $v$. After this initial state has been created, we then allocate our
priority queue $Q$, and set the current input ID $u$ to $0$.
Recall that the function $\taddrow{\Nt}{t}$ adds a new node to the node table $\Nt$
with birth time $t$, and returns the ID (i.e., index) of that new node in $\Nt$.
So, the operation $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ looks up the birth time
for input node ID $u$ in $\Nt_I$, adds a new node with this time to $\Nt_O$,
and stores the resulting (output) node ID in $v$.

To step back up through the pedigree, steps \algref{S1}--\algref{S8}
consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
In \algref{S1} we first find all edges in the input where the parent is equal to
our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
Then, in \algref{S2} we find all intersections between these input edges and the
mapped ancestry segments of their child nodes.
For each of these intersections we
create a new segment $y$ and insert it into the priority queue.
(In the paint pot analogy,
this stores in $Q$ all segments of color that must be transferred to the ancestor
with input ID $u$.)

After finding all overlaps and filling the priority queue with the resulting ancestry
segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
loop over the segments in the queue, considering each in turn until the queue is empty.
Step \algref{S3} builds a list $X$ of all segments covering the current left-most coordinate
(along with some bookkeeping to keep track of where our next
right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
this means that we have no overlapping ancestral segments at this coordinate and
we proceed to \algref{S4}.

In this case, only a single segment covers the leftmost region of the segments,
so there is no change in the mapping between input and output IDs,
and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
(The color in this segment of the offspring can simply be transferred to the parent.)
However, if other segments in $Q$ intersect with $x$
(since $x$ is the current left-most segment, these have left coordinate less then $\attrright{x}$),
then we can only transfer the portion of $x$ up until it overlaps with other segments.
Therefore, we change the left coordinate
of $x$ to this coordinate and reinsert it into $Q$ where it will be processed again
later, effectively cutting off (and propagating) the leftmost segment of $x$.
After doing this, we proceed to \algref{S7}, update the ancestry mapping for
the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.

On the other hand, if there is more than one segment in $X$, we know that an
overlap among the ancestral segments has occurred and we must update the output
nodes and edges accordingly. In step \algref{S5} we first allocate a new output
node $v$ for input node $u$, if it has not been done already (we may have many overlapping
segments along the genome). Step \algref{S6} then continues by creating a new
ancestry segment $\alpha$ mapping the current interval to the output node $v$.
We then iterate over all segments in $X$, adding the appropriate output edges
and editing and reinserting the ancestry segment $x$ into the priority queue if
necessary, as before.
